---
title: "Demo for flowmix shiny project"
author: "Henry He"
date: "2020/8/13"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=14, fig.height=8, echo=TRUE, eval=TRUE, cache=TRUE,
                      warning=FALSE, message=FALSE)
library(chron)
library(devtools)
library(dplyr)
library(tidyr)
library(data.table)
library(ggplot2)
library(gganimate)
library(lubridate)
library(plotly)
library(abind)
load_all('/home/shyun/repos/flowmix/flowmix') ## Replace with the location of on
                                              ## your computer
```


# Justin's comments for improving this demo

- Your code is very, very repetitious. Your code is not meant as a simple
  one-time project; we are writing code that is to be maintained by future
  collaborators of this project, for years to come. Please keep this in mind
  while you're writing code.
  
- While your demo app looked nice and polished, when I took a second look at
  this document, the app was not well documented at all. I spent a full day
  looking at this script and trying to improve it (and demonstrate the
  improvement, in terms of speed and readablility), but I'm not close to being
  done. This shows:
  
  1. How slow/inefficient your code was (e.g. many repeated `cbind`s and
  `rbind`s and `for(i in 1:100){ a = c(a, length(i))}`), and

  2. How difficult it is to read this document. (e.g. your code blocks were
  explained with partial sentences; at some point, you just wrote two incomplete
  sentences, not meant for anyone else but yourself:

```
- (single dataframe)
-  easier to handle using plot_ly() and highlight_key()
```
  
  3. a lot of copy pasting of code; for instance, column names were set by
  `colnames(pie_table)<-c("Time","cluster_1","cluster_2","cluster_3","cluster_4","cluster_5","cluster_6","cluster_7","cluster_8","cluster_9","cluster_10")`
  instead of `colnames(pie_table) <- c("Time", paste0("cluster_", 1:10))`.
  Surely, if you've taken coding class you are capable of the latter.

- You had some comments embedded in the code, which is great, but they were not
  comprehensive enough for me to know what was going on.

- I've mentioned this several times, but your code should be modularized into
  functions whenever possible. I can give you an example of (fake) readable code
  that I would have expected to see something like:
	 
```{r fake-but-clean-code, eval=FALSE}
filedir = "~/Downloads/for-henry/res.RDS"
res = readRDS(filedir)
pie_table = form_pie_table(res)
all_2d_tables = form_2d_tables(res)
plist = lapply(all_2d_tables, plot_2d_table())
subplot(plist[[1]], plist[[2]], plist[[3]],
        shareX = TRUE, shareY = TRUE)
```

Then, if I want to know how `form_pie_table()` works, I can simply look into the
function definition, something like:

```{r, more-fake-but-clean-code, eval=FALSE}
##' Function to form a table of cluster probabilities.
##' 
##' @param res Object of class "flowmix"
##'
##' @return Matrix of size (TT x 5), whose columns are ..
form_pie_table <- function(res){
  res
}
```

- The long for-loop for producing the three tables
  (e.g. `diam_mid_chl_small_table`) should be rewritten to be more efficient and
  readable. I would start with the following steps (1-4):
  1. First prescribe to yourself what the end product needs to be e.g. three
     tables each containing count data from two axes at a time, as well as the x
     and y values of the points to plot.
  2. Make and call a function for parts of the code that are repeated. A lot of
     code is simply copy-pasted here (the whole code block starting with ~##
     Temporary~ exists three times) -- you should avoid this.
     + You can make a separate file called "shiny-helpers.R" and source it in.
     + I'd recommend that you make the helper functions well documented; if you
       want to see an example, just look at the `flowmix` R package, under like
       the `objective()` function here
       https://github.com/sangwon-hyun/flowmix/blob/master/flowmix/R/objective.R
  3. Try to eliminate as many duplicated operations as you can. You've clearly
     copy-pasted identical code multiple times. You should avoid this.
  4. Try to avoid the repeated concatenations e.g. `diam_mid_chl_small_table <-
     rbind(diam_mid_chl_small_table,y)` This makes the code very slow.

- (I've in fact done a lot of this already; see my new code block labeled
  "2d-tables" which replaces your long for-loop. I've used a function called
  `collapse_3d_to_2d(y, counts, dims=1:2)` from my flowmix R package, just so
  you know.

- As much as possible, don't hard-code anything (use `numclust` instead of `10`
  for number of clusters, `dimdat` instead of `3` for dimension of data, `TT`
  instead of 296, etc.)

- In the following block of code, instead of the loop, you could just do this:
 `time = sapply(ylist, nrow)`. This is faster and shorter. (I understand why you
 might do this if you are coming from a C background, but in R, for-loops +
 concatenating is **very slow**, since R has to make new copies and allocate new
 memory of the object for every iteration of the for loop; see
 https://stackoverflow.com/questions/14693956/how-can-i-prevent-rbind-from-geting-really-slow-as-dataframe-grows-larger
 for an entertaining discussion about this).

```{r, eval=FALSE}
time <- c()
for(i in ylist){
  time <- c(time,nrow(i))
}
time_column <- rep(Time, time)
```

- Your object names should be more descriptive. You have `time` and `Times` and
  `TT`; it's not clear which is what. In fact, `time` is the number of rows of
  each element in `ylist`, and not related to time at all.

- Just to be clear, in all of my code, I'm using `TT` to encode the *number* of
  time points (e.g. `TT == length(ylist)`), and `tt` to mean a specific time
  point e.g. `for(tt in 1:TT){ ...}`. It would be great if we could keep this a
  convention.
   

- Now that you know how to program in R, you should definitely read this style
  guide http://adv-r.had.co.nz/Style.html (old but short and useful version) and
  https://style.tidyverse.org/syntax.html (newer and longer version) and adhere
  rigorously. For future reference, Hadley Wickham is a statistician who is
  famous for improving statistical software (e.g. R, tidyverse, dplyr); his
  advice found on the web is usually sound and useful.
   
- This document is clearly not knittable. That's, in general, the whole point of
  using R Markdown -- you have a single ".Rmd" file that you can knit to
  **reproduce** your data analysis. I should be able to hit "knit"
  (Ctrl+Shift+K) in my Rstudio and be able to produce a complete document; the
  same one as yours. From hereon, when you share Rmd documents, could you make
  sure that these are completely knittable.

- The code doesn't work as intended -- my new code can't reproduce your `p1` --
  which is the obvious first thing to fix.

- If you look at the code block called `animation1`, it seems that the
  background scatterplot points (not the cluster means) are not designed to move
  over time (I could be wrong -- let me know if I am). You only use the first
  500 columns of the large data frame. What happens when you use more rows, or
  all rows? Is this too slow, or prohibitive memory-wise? If so, doesn't that
  mean that this approach either needs to be improved or replaced?

The rest of this document (up till before "The rest of the code") is my attempt
at cleaning while trying to understand what you did. Please study it carefully,
compare it to your old code (feel free to ask any questions), address all of the
points I've made, and **make a more complete demo.Rmd** and share it with me by
the next time we meet.


# Load Data

The file res.RDS will be loaded into a list called `res`.

```{r}
## filedir = "C:/Users/Administrator/Documents/GitHub/flowcy-shiny/learning-plotly/res.RDS"
filedir = "~/Downloads/for-henry/res.RDS"
res = readRDS(filedir)
countslist = res$countslist
ylist = res$ylist
mn = res$mn
numclust = res$numclust
TT = length(ylist)
Time <- names(ylist)
```

# Read data

Some points:

1. Elements from `res` will stored into three dataframes.
2. There are 296 unique time points in the data.
3. The response (`ylist`) will be collapsed into two dimensions at a time, and
   will be stored into 3 dataframes called `diam_chl_table`, `pe_chl_table`, and
   `diam_pe_table`.

First, setting up the data:

```{r pie_table}
## Form `pie_table`
pie_table <- data.frame(res$pie)
pie_table <- setDT(pie_table, keep.rownames = "Time")[]
colnames(pie_table) <- c("Time", paste0("cluster_", 1:10))

# Reshape pie_table into a dataframe with 3 columns：
# Time : time corresponding to this row
# Cluster : Which cluster is it; an integer from 1 to 10
# Prob : the probability of this cluster at this time
pie_table <- pie_table%>%
  gather(var,val,-c(1))%>%
  separate(var,c("type","cluster"))%>%
  .[-2]
colnames(pie_table) <- c("Time", "Cluster","Prob")

## Now, combining the **cluster mean** information with the **cluster
## probability**, in `pie_table`. (Every row of `pie_table` is a unique (time,
## cluster, point); SH: is this correct?)
dfs = lapply(1:numclust, function(iclust){
  one_table = tibble(diam_mid_pie = mn[,1,iclust],
                     chl_small_pie = mn[,1,iclust],
                     pe_pie = mn[,1,iclust],
                     Time = Time, Cluster = rep(iclust, TT))
  colnames(one_table)[1:3] <- c("diam_mid_pie","chl_small_pie","pe_pie")
  one_table
})
mn_table = bind_rows(dfs)
pie_table = merge(pie_table, mn_table)
```


```{r y_table}
## Form a y_table
y_table = do.call(rbind, ylist) %>% as.data.frame()
reptimes = sapply(ylist, nrow)
time_column <- rep(Time, reptimes)
y_table$Time <- time_column
```

```{r 2d-tables, eval=TRUE}
time_justin_code = microbenchmark::microbenchmark({

  ## Setup
  alltimes = names(ylist)
  TT = length(ylist)
  tablist = list()
  dimslist = list(c(1,2), c(2,3), c(3,1))

  ## Obtain all the tables once.
  for(ii in 1:length(dimslist)){

    dims = dimslist[[ii]]

    ## Form the table
    all_2d_tables = lapply(1:TT, function(tt){
      y = ylist[[tt]][,dims]
      counts = countslist[[tt]]
      collapse_3d_to_2d(y, counts, dims=1:2)
    })
    combined_2d_table = do.call(rbind, all_2d_tables) %>% as_tibble()

    ## Add time
    reptimes = sapply(all_2d_tables, nrow)
    times = rep(alltimes, reptimes)
    combined_2d_table[,"times"] = times
    newname = paste0("counts_", ii)
    combined_2d_table <- rename(combined_2d_table, !!newname:=counts)
    tablist[[ii]] = combined_2d_table
  }
  diam_chl_table <- tablist[[1]]
  pe_chl_table <- tablist[[2]]
  diam_pe_table <- tablist[[3]] 
}, times = 1)

## The above code is 5-6 times faster than your old code, and is easier to read.

##       min        lq      mean    median        uq       max neval
##  5.837105  5.837105  5.837105  5.837105  5.837105  5.837105     1
## 32.042188 32.042188 32.042188 32.042188 32.042188 32.042188     1
```

# Animation of three scatterplots 

- (single dataframe)
-  easier to handle using plot_ly() and highlight_key()

```{r animation1}
# To have a quickier view; Delete this line on final presentation
# generate three plots

# Highlight key 
key <- highlight_key(pie_table,~Cluster)

p1 <- plot_ly(diam_chl_table[1:500,],
              x = ~diam_mid,
              y = ~chl_small,
              size = ~ counts_1,
              type = 'scatter',
              mode = 'markers',sizes = c(0,1000), opacity = 0.5,alpha = 0.3,
              showlegend = FALSE) %>%
  add_trace(data = key,
            x = ~diam_mid_pie,
            y = ~chl_small_pie,
            frame = ~Time,
            size = 5,
            hoverinfo = 'text',
            text = ~paste("Cluster", Cluster),
            hovertext = ~paste("Cluster", Cluster),
            textposition = "bottom center",
            mode = 'markers+text',
            textfont = list(size = 14),
            opacity = 0.9)
p1
```

# The rest of the code

## p2 and p3

```{r, eval=FALSE}

p2 <- plot_ly(pe_chl_small_table[1:500,],
              y = ~pe,
              x = ~chl,
              size = ~ counts_2,
              type = 'scatter', mode = 'markers',alpha = 0.3,
              opacity=0.5,sizes = c(0,1000),showlegend = FALSE)%>%
  add_trace(data = key,
            x = ~chl_small_pie,
            y = ~pe_pie,
            frame = ~Time,
            showlegend = FALSE, 
            size = 5,
            hoverinfo = 'text',
            text = ~paste("Cluster",Cluster),
            hovertext = ~paste("Cluster",Cluster),
            textposition = "bottom center",
            mode = 'markers+text',
            textfont = list(size = 14),
            opacity = 0.9
            )

p3<-plot_ly(diam_mid_pe_table[1:500,],
            x= ~pe,
            y = ~diam_mid,
            size = ~ counts_3,
            type = 'scatter',
            mode = 'markers',
            opacity=0.5, sizes = c(0,1000),alpha = 0.3, showlegend = FALSE)%>%
  add_trace(data = key,
            x = ~pe_pie,
            y = ~diam_mid_pie,
            frame = ~Time,
            size = 5,
            hoverinfo = 'text',
            text = ~paste("Cluster",Cluster),
            hovertext = ~paste("Cluster",Cluster),
            textposition = "bottom center",
            mode = 'markers+text',
            textfont = list(size = 14),
            opacity = 0.9
            )
scatterplot <- subplot(p1,p2,p3,shareX = TRUE,shareY = TRUE)
scatterplot
```

## Animation of the covariates table 
```{r, eval=FALSE}

# Colors
myColor = c("red","green","orange","blue","brown","deeppink","navyblue","orchid","seagreen","gold")


# create highlight key, which highlights by cluster
key <- highlight_key(pie_table,~Cluster)
pie_plot <- plot_ly(key)%>%
  group_by(Cluster)%>%
  add_lines(x = ~Time , y = ~Prob, color = ~Cluster, colors = myColor ,)%>%
  add_segments(x = ~Time, xend = ~Time, y = 0, yend = 1, frame = ~Time,showlegend = FALSE, color = "red",showlegend = FALSE)

# modify the x-axis  
pie_plot <- pie_plot%>%layout(
  xaxis = list(
    type = "date",
    tickformat = "%d%M%Y",showticklabels = FALSE)
)
pie_plot






```

Combine two plots
```{r, eval=FALSE}
# generate three plots
 subplot(scatterplot,pie_plot,nrows = 2)
```

