---
title: "Learning plotly and linked animations"
author: Sangwon Hyun
date: Jul 23, 2020
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=14, fig.height=8, echo=TRUE, eval=TRUE, cache=TRUE,
                      warning=FALSE, message=FALSE)
## la('flowcy')
figdir = "~/Dropbox/research/usc/flow-cytometry/figures"
library(gridExtra)
library(tidyverse)
library(dplyr)
```


## Goal

Learn about linked animations in R + plotly.

## Main reference

https://plotly-r.com/client-side-linking.html#linking-animated-views

- "This section focuses on a particular approach to *linking views* known as
graphical (database) queries using the R package plotly."

- "With plotly, one can write R code to pose graphical queries that operate
entirely client-side in a web browser (i.e., no special web server or callback
to R is required)."

- "In addition to teaching you how to pose queries with the `highlight_key()`
function, this section shows you how to control how queries are triggered and
visually rendered via the `highlight()` function."

*(Emphasis and formatting mine).*

## First example

Take a data frame called `mtcars`.

```{r}
mtcars %>% head() %>% print()
```

The option `highlight_key(~cyl)` lets you hover to a data point, and highlight
  the points with the same `cyl` values.

Use `highlight()` to customize the various interactions e.g. hover, single
  click, double click.

In the background, `highlight(on="plotly_hover")` performs a SQL query of the
  form: `SELECT * FROM mtcars WHERE cyl IN $SELECTION_VALUE`.

```{r, echo=TRUE, eval=TRUE, fig.width=10, fig.height = 5}
library(plotly)
mtcars %>%
  highlight_key(~cyl) %>%
  plot_ly(
    x = ~wt, y = ~mpg, text = ~cyl, mode = "markers+text", 
    textposition = "top", hoverinfo = "x+y"
  ) %>%
  highlight(on = "plotly_hover", off = "plotly_doubleclick")
```

## More involved example

Now, load the `txhousing` dataset, whose rows are *monthly median sales* of real estate in regions of Texa:
```{r}
data(txhousing, package = "ggplot2")
txhousing %>% head() %>% print()
```

Now, we'll create an interactive map:

First, declare `city` as the SQL 'query by' column. *THIS IS THE ONLY
INTERACTIVE ELEMENT IN THIS EXAMPLE*.

```{r}
tx <- highlight_key(txhousing, ~city)
```

Then, initiate a plotly object called `base`, then pipe it through some `plotly` functions to add features.
```{r}
# initiate a plotly object
base <- plot_ly(tx, color = I("black")) %>% group_by(city)
base %>%
  group_by(city) %>%
  add_lines(x = ~date, y = ~median) -> time_series
```

## Widget for selecting certain rows of data

Querying a city via direct manipulation is somewhat helpful for focusing on a
particular time series, but it’s not so helpful for 
- querying a city by name,
- and/or comparing multiple cities at once. 

As it turns out, plotly makes it easy to add a `selectize.js` powered dropdown
widget for querying by name (aka indirect manipulation) by setting 
`selectize =TRUE`.

When it comes to comparing multiple cities, we want to be able to both 
- retain previous selections (`persistent = TRUE`), as well as 
- control the highlighting color (`dynamic = TRUE`).

<!-- This videos explains how to use these features in -->
<!-- Figure 16.3 to compare pricing across different cities. -->
Try this out:

```{r}
time_series %>% 
highlight(
  on = "plotly_click", 
  selectize = TRUE, 
  dynamic = TRUE, 
  persistent = TRUE
) 
```

## Linking

Now, (finally) we get to linking.

One obvious thing from playing around with the above application is that not
every city has complete pricing information (e.g., South Padre Island, San
Marcos, etc).

To learn more about what cities are missing information as well as how that
missingness is structured, we aim to link 
- a view of the raw time series 
- to a dot-plot of the corresponding number of missing values per city.

In addition to making it easy to see how cities rank in terms of missing house
prices, it also provides a way to query the corresponding time series (i.e.,
reveal the structure of those missing values) by brushing cities in the
dot-plot.

This general pattern of linking aggregated views of the data to more detailed
views fits the famous and practical information visualization advice from
Shneiderman (1996): “Overview first, zoom and filter, then details on demand”.

```{r eval=TRUE}
data(txhousing, package = "ggplot2")
tx <- highlight_key(txhousing, ~city)
base <- plot_ly(tx, color = I("black")) %>%
  group_by(city)

time_series <- base %>%
  group_by(city) %>%
  add_lines(x = ~date, y = ~median)


dot_plot <- base %>%
  summarise(miss = sum(is.na(median))) %>%
  filter(miss > 0) %>%
  add_markers(
    x = ~miss, 
    y = ~forcats::fct_reorder(city, miss), 
    hoverinfo = "x+y") %>%
  layout(
    xaxis = list(title = "Number of months missing"),
    yaxis = list(title = ""))  %>% 
  highlight(on = "plotly_click", dynamic = TRUE, selectize = TRUE)


dot_plot
time_series

subplot(dot_plot,
        time_series,
        widths = c(.2, .8), titleX = TRUE) %>%
  layout(showlegend = FALSE) %>%
  highlight(on = "plotly_selected", dynamic = TRUE, selectize = TRUE)
```

(I can't get these to link; this is maybe because the documentation does not
specify what `time_series` is, so I'm just guessing.)

Some takeaways
- only one `highlight()`  is possible for each visualization.
- `subplot(a,b)` links two plotly plots `a` and `b`. `layout()` is formatting
  for the subplots.

**Q**: I can't figure out what the background grey is doing.

How does plotly know to highlight the time series when markers in the dot-plot
are selected? The answer lies in what data values are embedded in the graphical
markers via `highlight_key()`.

What happens in the background? First imagine a linked database query being
performed behind the scenes:

![Figure 16.5](./pipeline.svg)

When ‘South Padre Island’ is selected, it first filters the aggregated dot-plot
data down to just that one row, then it filters down the raw time-series data
down to every row with ‘South Padre Island’ as a city.

**The drawing logic will then call `Plotly.addTrace()` with the newly filtered
data** which adds a new **graphical layer** representing the selection, allowing
us to have finely-tuned control over the visual encoding of the data query.


The biggest advantage of drawing an entirely new **graphical layer** with the
filtered data is that it becomes easy to leverage *statistical trace types* for
producing summaries that are conditional on the query; here is another example
of doing this with histograms:

```{r}
hist <- add_histogram(
  base,
  x = ~median, 
  histnorm = "probability density"
)

subplot(time_series, hist, nrows = 2) %>%
  layout(barmode = "overlay", showlegend = FALSE) %>%
  highlight(
    dynamic = TRUE, 
    selectize = TRUE, 
    selected = attrs_selected(opacity = 0.3)
  )
```

## Highlight vs Filter

A **highlight** event dims the opacity of existing marks, then adds an additional
graphical layer representing the selection.

A **filter** event completely remove existing marks and rescales axes to the
remaining data.

(Filtering uses the `crosstalk` R library.)

```{r}
library(crosstalk)

# generally speaking, use a "unique" key for filter, 
# especially when you have multiple filters!
tx <- highlight_key(txhousing)

gg <- ggplot(tx) + geom_line(aes(date, median, group = city))

## 1. FILTER using filter_select() + bscols()
filter <- bscols(
  filter_select("id", "Select a city", tx, ~city), ## A DIRECT
  ggplotly(gg, dynamicTicks = TRUE),
  widths = c(12, 12)
)
filter

## 2. HIGHLIGHT using highlight()
tx2 <- highlight_key(txhousing, ~city, "Select a city")
gg <- ggplot(tx2) + geom_line(aes(date, median, group = city))
select <- highlight(
  ggplotly(gg, tooltip = "city"), 
  selectize = TRUE, persistent = TRUE
)

select
```

Some takeaways:

- `bscols()` is for making side-by-side HTML elements. *This will be important*
- The `highlight` functionality is from the plotly side; the `filter_select`
  functionality heavily borrows from `crosstalk`.


## Multiple widgets

Show only a subset of years or cities, and also only show medians with `sales`
in some range:

```{r}
library(crosstalk) 
tx <- highlight_key(txhousing)
widgets <- bscols(
  widths = c(12, 12, 12),
  filter_select("city", "Cities", tx, ~city),
  filter_slider("sales", "Sales", tx, ~sales),
  filter_checkbox("year", "Years", tx, ~year, inline = TRUE)
)
bscols(
  widths = c(4, 8), widgets, 
  plot_ly(tx, x = ~date, y = ~median, showlegend = FALSE) %>% 
    add_lines(color = ~city, colors = "black")
)
```


As Figure 16.9 demonstrates, filter and highlight events can work in conjunction
with various htmlwidgets. In fact, since the semantics of filter are more
well-defined than highlight, linking filter events across htmlwidgets via
crosstalk should generally be more well-supported.

```{r}
eqs <- highlight_key(quakes)
stations <- filter_slider(
  "station", "Number of Stations", 
  eqs, ~stations
)

p <- plot_ly(eqs, x = ~depth, y = ~mag) %>% 
  add_markers(alpha = 0.5) %>% 
  highlight("plotly_selected")

library(leaflet)
map <- leaflet(eqs) %>% 
  addTiles() %>% 
  addCircles()

bscols(
  widths = c(6, 6, 3), 
  p, map, stations
)
```

## Some thoughts

I'm guessing we will have to make a choice (this is really a luxury) between the
widget functionality that Shiny provides, and what we can do directly in
conjunction with plotly. The last paragraph of chapter 16.5 says this:

"**16.5 Limitations**

The graphical querying framework presented here is for posing database queries
between multiple graphs via direct manipulation. For serious statistical
analysis, one often needs to link other data views (i.e., text-based summaries,
tables, etc) in other arbitrary ways. For these use cases, the R package shiny
makes it very easy to build on concepts we’ve already covered to build more
powerful client-server applications entirely in R, without having to learn any
HTML, CSS, or JavaScript. The next Chapter 17 gives a brief introduction to
shiny, then dives right into concepts related to linking plotly graphics to
other arbitrary views."

So, it seems to me that both chapters 16 *and* 17 are requisite readings.

## Finally, Animations

"The graphical querying framework (Section 16.1) works in tandem with key-frame
animations Section (14)."


```{r}
library(gapminder)
gapminder %>% print()
```

The dataset contains rows equal to unique country-years. 

The features are `continent`, `lifeExp`, `pop`, and `gdpPercap`.

The following animation does a few things:
- Draws a scatter plot over time `frame=year`, of `gdpPercap` and `lifeExp`.
- Colors by continent `color=continent`.
- Highlights the animation by continent `g <- highlight_key(gapminder,
  ~continent)`.
  + Hovering is enabled by `highlight(..., "plotly_hover")`.
  + If you put an 
- Adds a best-fit line `geom_smooth(method='lm')` between `gdpPercap` and
  `lifeExp`.

```{r}
g <- highlight_key(gapminder, ~continent)
gg <- ggplot(g, aes(gdpPercap, lifeExp, color = continent, frame = year)) +
  geom_point(aes(size = pop, ids = country)) +
  geom_smooth(se = FALSE, method = "lm") +
  scale_x_log10()
highlight(ggplotly(gg), "plotly_hover")
```


## Linked Animations

(We first need to install an R package from github using `Rscript -e
"devtools::install_github('cpsievert/plotly_book')"`.)

Now, let's try to link two animations.

```{r}
## data(gap, package = "plotlyBook")
load("./gap.rda") ## from https://github.com/cpsievert/plotly_book/blob/master/data/gap.rda
gap %>% print()
```

The dataset is a table whose rows are country-years, and contains features such
as life expectancy, gdp per capita, area and population density.


The following animation does a few things:

- Highlighting key is by country `gapKey <- highlight_key(gap, ~country)`.

- Basically, these are two different representations of one data frame; each row
  is one data point in both plots, but the x and y axes are different.
  - **p1** Draws a scatterplot of country vs population density `plot_ly(gap, y
    = ~country, x = ~popDen, hoverinfo = "x")`
  - **p2** Draws a scatterplot between per capita gdp vs life expectancy, with points sized by  `popDen`.
  
- Animation is over years, since `frame=~year` in both **p1** and **p2**

```{r}
gapKey <- highlight_key(gap, ~country)

p1 <- plot_ly(gap, y = ~country, x = ~popDen, hoverinfo = "x") %>%
  add_markers(alpha = 0.1, color = I("black")) %>%
  add_markers(
    data = gapKey, 
    frame = ~year, 
    ids = ~country, 
    color = I("red")
  ) %>%
  layout(xaxis = list(type = "log"))

p2 <- plot_ly(gap, x = ~gdpPercap, y = ~lifeExp, size = ~popDen, 
              text = ~country, hoverinfo = "text") %>%
  add_markers(color = I("black"), alpha = 0.1) %>%
  add_markers(
    data = gapKey, 
    frame = ~year, 
    ids = ~country, 
    color = I("red")
  ) %>%
  layout(xaxis = list(type = "log"))

subplot(p1, p2, nrows = 1, widths = c(0.3, 0.7), titleX = TRUE) %>%
  hide_legend() %>%
  animation_opts(1000, redraw = FALSE) %>%
  layout(hovermode = "y", margin = list(l = 100)) %>%
  highlight(
    "plotly_selected", 
    color = "blue", 
    opacityDim = 1, 
    hoverinfo = "none"
  )
```

## What's next?

Next steps:

- Try out the examples in 16.4 (in a similar fashion/rigor as above).
- Go over chapter 17.
- Read frame-wise animations Section (14).

Some questions to answer (**SH:** Henry please follow up with me on these points.)

- Can we link animations coming from more than one data frame? or do we need to
  somehow make a giant data frame.
- How to use these tools for visualizing our data?
- How to use widgets with `subplots()`?
- How to host in a Shiny app?
- What can we use from native shiny, and what can we use from plotly
  (i.e. building widgets entirely using plotly).


## An example using our data

Load our data.

```{r}
filedir = "./res.RDS"
res = readRDS(filedir)
```

Make some scatterplots:

```{r, fig.width=20, fig.height=7}
## Summarize the ylist by collapsing the counts into two dimensions
devtools::load_all('~/repos/flowcy/flowcy')

## Use the functionality in the "flowcy" R package.
par(mfrow = c(1, 3))
tt = 50
library(dplyr)
for(dims in list(c(1,2), c(2,3), c(3,1))){
  one_dim_scatterplot(res$ylist, obj=NULL, tt=tt,
                      countslist = res$countslist, cex_fac=20,
                      dims = dims)
}
```

Now, make an animation:
```{r}
times = 200:296
fac = 2

## First pair of dimensions
dims = c(1:2)
ylist_summary <- Map(collapse_3d_to_2d, res$ylist, res$countslist,
                     dims = rep(list(dims), length(res$ylist)))[times]
ylist_summary <- lapply(ylist_summary, as_tibble)
ylist_summary <- Map(function(dat, nm){cbind(dat, time=nm)},
                     ylist_summary, names(ylist_summary))
ymat = do.call(rbind, ylist_summary)
ymat$counts = ymat$counts/max(ymat$counts) %>% sqrt() * fac
p1 = plot_ly(ymat, x= ~diam_mid,y = ~chl_small,
             marker = list(size = ~counts, opacity=0.3),
             frame = ~time,
             col="blue")
## p1 %>% animation_opts(50, redraw = FALSE)


## Second pair of dimensions
dims = c(2:3)
ylist_summary <- Map(collapse_3d_to_2d, res$ylist, res$countslist,
                     dims = rep(list(dims), length(res$ylist)))[times]
ylist_summary <- lapply(ylist_summary, as_tibble)
ylist_summary <- Map(function(dat, nm){cbind(dat, time=nm)},
                     ylist_summary, names(ylist_summary))
ymat = do.call(rbind, ylist_summary)
ymat$counts = ymat$counts/max(ymat$counts) %>% sqrt() * fac
p2 = plot_ly(ymat, x= ~chl_small,y = ~pe,
             marker = list(size = ~counts, opacity=0.3),
             frame = ~time,
             col="blue")

## Third pair of dimensions
dims = c(3,1)
ylist_summary <- Map(collapse_3d_to_2d, res$ylist, res$countslist,
                     dims = rep(list(dims), length(res$ylist)))[times]
ylist_summary <- lapply(ylist_summary, as_tibble)
ylist_summary <- Map(function(dat, nm){cbind(dat, time=nm)},
                     ylist_summary, names(ylist_summary))
ymat = do.call(rbind, ylist_summary)
ymat$counts = ymat$counts/max(ymat$counts) %>% sqrt() * fac
p3 = plot_ly(ymat, x= ~pe,y = ~diam_mid,
             marker = list(size = ~counts, opacity=0.3),
             frame = ~time,
             col="blue")

subplot(p1, p2, p3) %>% animation_opts(50, redraw = FALSE) %>% layout(width=1500, height=500)
```

Also plot the covariates:
```{r}
X = res$X
X = cbind(time=rownames(X),X)
db = X[,] %>% as_tibble() %>% reshape2::melt(id.vars=c("time")) 
## '%ni%' <- Negate('%in%')
## db$variable %ni% c("par","sst", "sss", "nitrate_WOA_clim")
colors = rep("grey", ncol(X)-1)
colors[colnames(X) %in% c("par","sst", "sss", "nitrate_WOA_clim")] = RColorBrewer::brewer.pal(4, "Set3")
width = rep(1, ncol(X))
width[colnames(X) %in% c("par","sst", "sss", "nitrate_WOA_clim")] = 10
p4 = db %>% plot_ly(color = ~variable, x = ~time, y = ~value,
                    colors = colors, width=2000, height=400
                    ## line = list(width = width),
                    ) %>% add_lines()
p4
```

## What's next? 

(**SH:** Henry please follow up with me on these points.)

- Link *three* of these scatterplots (`p1`, `p2`, and `p3`, and a covariate plot`p4`).
- The three axes are all different; fix them over time.
- The point sizes are weird; make sure the plotly point sizes are consistent
  with the original plots.
- Make it a joint animation.
- Code blocks creating `p1`, `p2`, and `p3` are messy; clean up and remove
  redundancy.
- Add means (red dots) and probabilities (red dot sizes).
- Add ellipses (lower priority for now).
- For the covariate plot in `p4`, put the grey covariates in the background, and
  bring the colored covariates (par, sss, sst, nitrate) to the front.
- Interactivity: when a mouse click or hover occurs over one of the red dots;
  the red means in *all three plots* corresponding to that cluster (e.g. cluster
  number 3) should be highlighted.
- Do we want to use `ggplotly()` or `plot_ly()`? Reading
  https://plotly-r.com/animating-views.html could be helpful.
- In `p4`, add vertical line that moves as a slider moves.
- Slider should just have a few ticks.
- The animation might be faster if we made it depend the same single data frame?
  (This is just a hunch; it seems that animations might be produced by issuing
  separte SQL queries to data matrices, every time the slider value
  changes. This can be slow).
- Smoother animation of scatter plots?
- Any other suggestions?
- (Longer term) All of this code should be organized as modular functions and
  the code and functions should all be well documented.
- Generally speaking, pro-actively do reading and document (i.e. take notes of)
  what you've learned!
- Find out how to host this in a Shiny app.
- After we fully figure out how to the 3 scatter plots + 1 covariate plot *and*
  how to host it, we should add covariate tables via hovering.
