---
title: "Shiny app demo"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 7, echo = TRUE, warning = FALSE,
                      message = FALSE, eval = TRUE, cache = FALSE)
```

# Setup

Load libraries.

```{r}
library(shiny)
library(plotly)
library(dplyr)
library(plotly)
library(chron) ## SH: why do we need this? If we don't, delte it.
library(devtools)
library(tidyr)
library(data.table)
library(ggplot2)
library(gganimate)
library(randomcoloR)
source("viz.R")
```

# Notes from Justin

(You can erase each item after it has been taken care of.)

- Your app should be a much simpler call to the shiny. The call to the app
  should be `viz.flowmix()`.

- Your code is not organized enough. Your code should all be functions that are
  documented to completion, with `@param` and `@return` tags, as well as a
  description of what the function does.

- Your code shouldn't rely on external calls to `make_prob_table()` or
  `make_prob_plot` or `make_covariates_table()` or
  `make_covariates_plot()`. This should all be contained in `viz.flowmix()`.

- Your app doesn't work. `covariates_plot` doesn't show me anything. Anything
  after that doesn't work either.

- `convert_ylist_2d()` depends on `collapse_3d_to_2d()` function. You don't
     have this function in the github package. It seems to exist in
     `demo-wip/improved_demo.Rmd` but seeing that.

- `convert_ylist_2d()` is too specific for our data. It assumes that the three
     dimensions of data are named diam, pe and chl; it has variables called
     `diam_chl_table` `pe_chl_table` and `diam_pe_table`.

- Moving forward, your code should not assume that any of the data has any
  specific column names. It is *general* data; for `T` time points in total,
  `ylist` is a list of $n_t$ by $d$ matrices, where $n_t$ is the number of data
  points at time $t$. The $p$ covariates are a single matrix of dimension $T
  \times p$.

- You should clean the github repository, and keep it clean from now on. All of
  your functions are in `viz.R`. Right now, you have `app.R` and `v0.Rmd` and
  `shiny-app-demo.R`, as well as several folders. I'm going to clean it once, so
  that there are only two demo files:
  + `shiny-app-demo.Rmd`, which contains the latest code.
  + `plotly-demo.Rmd`, which contains the latest code for the demo *before* we
    started using Shiny; this is the code Jacob and I saw when you present last
    time.

- I'm continuously mentioning this, but it's important to make sure your demo
  files are error-free. What I mean by error-free is this: in an R file, you
  make sure that I can open up R, and run the code from beginning to end, and
  produce the same output or app as you, with no bugs or errors. (I've never
  seen an error-free demo script from you yet). Likewise for an Rmd file; I
  think your workflow is to just run the code within an Rmd file, but that often
  results in errors in the code. An rmd file being error-free means that you're
  able to knit the file (Ctrl-Shift-K). If you don't make the scripts
  error-free, there is nothing I can help with (I spent quite a few hours today
  just cleaning your code, just to realize that the code doesn't work.)

- For the two Rmd files, you should knit to produce the html files, and push
  them to github. Just to be clear, you should always have the latest version of
  `plotly-demo.Rmd` and `plotly-demo.html`.

- I've tried to clean the functions `viz.R`, but haven't been able to finish.
  + Everything in this file should be function definitions. No calls to R
    libraries or loading data files.
  + Whenever a function is called, it should either be defined within the same
    file, or you should prepend the function call with a tag like
    `packagename::`. For instance, if you call `collapse_3d_to_2d()` from the
    flowmix R package, you should write `flowmix::collapse_3d_to_2d()`.
  + You should look at all the comments that start with "SH:", and address them.
  

- This is a good time for you to take a look at the flowmix R package here, to
  familarize yourself with the data and model:
  https://github.com/sangwon-hyun/flowmix
  + In particular, you should try running the demo file:
    https://github.com/sangwon-hyun/flowmix/blob/master/demo.Rmd
  + The example output is here:
  + https://htmlpreview.github.io/?https://github.com/sangwon-hyun/flowmix/blob/master/demo.html


# Main code

Load data.

```{r}
## filedir = "C:/Users/heqin/OneDrive/Desktop/for-henry (2)/res.RDS"
datadir = "~/repos/flowmix-shiny"
res = readRDS(file = file.path(datadir, "res.RDS"))
res$prob = res$pie
```

Make cluster probability plotly object.

```{r}
prob_table = res %>% make_prob_table()
prob_plot <- prob_table %>% make_prob_plot()
```

Make covariates plotly object:
```{r, eval=FALSE}
covariates <- res %>% make_covariates_table()
covariates_plot <- covariates %>% make_covariates_plot()
covariates_plot
```

SH: I realized this doesn't work. So the remaining code doesn't work
either. I've set `eval=FALSE`.
 

```{r, eval=FALSE}
ui <- fluidPage(
    fluidRow(
        column(6,
               plotlyOutput(outputId = "row_selected")),
        column(4,
               div(DT::dataTableOutput("table"), style = "font-size:60%; width: 10%"))      
    ),
    fluidRow(
        column(6,
               plotlyOutput("p") 
               ),
        column(4,
               verbatimTextOutput("event")
        )
        
    )
)

server <- function(input, output, session) {
    output$event <- renderPrint({
      # get click information
        d <- event_data("plotly_click")
        if(is.null(d))
            "The selected cluster number is Here"
        else{
            d$key
        }
        
    })
    output$table = DT::renderDataTable({
        beta_index_first = 0
        beta_index_second = 0
        selected = event_data("plotly_click")
        if(!is.null(selected)){
            cluster_num = as.numeric(selected$key)
            alpha = t(round(as.data.frame(res$alpha)[cluster_num,],2))
            beta_index_first = (cluster_num-1)*3 + 1
            beta_index_second = beta_index_first + 2
            if(beta_index_first != 0){
                beta = round(as.data.frame(res$beta)[,beta_index_first:beta_index_second],2)
            }
            cbind(alpha,beta)
        }
    })
    output$row_selected <- renderPlotly({
       subplot(cv_plot,pie_plot,nrows = 2)
    })
    
    observeEvent(input$table_rows_selected, {
      s = input$table_rows_selected
      covariates = res$X[,s]
      covariates = as.data.frame(covariates)
      covariates = setDT(covariates, keep.rownames = "Time")[]
      covariates <- covariates%>%
        gather(var,val,-c(1))
      colnames(covariates) = c("Time","Trace","Value")
      plotlyProxy("row_selected",session)%>%
        plotlyProxyInvoke("addTraces",list(x = covariates$Time,y = covariates$Value, colors = "red")
          
        )
      
    })
    
}
```

Run the shiny app.

```[r, eval=FALSE}
# Run the application
shinyApp(ui = ui, server = server)
```
